---
title: "Intro to Shiny apps"
output: 
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
date: "2023-03-02"
runtime: shiny
author: Ezra Morrison (<demorrison@ucdavis.edu>)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Shiny in RMarkdown

We can create Shiny apps directly from .Rmd files; just add `runtime: shiny` to the header, e.g.,

```
---
title: "Shiny in RMarkdown"
output: html_document
date: "2023-02-27"
runtime: shiny
---
```

Then we can add shiny input widgets and use their values to inform our analyses:

```{r pressure, echo=TRUE}
library(shiny)

sliderInput(
  inputId = "bins", 
  label = "how many bins?", 
  min = 1, 
  max = 50, 
  val = 10)

renderPlot( # this wrapper tells Shiny to update the histogram in response to user actions
  hist(iris$Sepal.Length, breaks = input$bins)) # We reference the values of Shiny inputs with the prefix `input$`

```

# UI-server structure

We can build more complex shiny apps using a server and UI structure. Copy-paste the following code block into the R console and you'll get a basic app: 

```r

library(shiny)

# next two lines optional, for debugging:
library("reactlog")
reactlog_enable()
# if you include these lines, you can press control-F3 (windows) or command-F3 (mac) while app is running to get an interactive graph of your app's internal structure

# Define UI for application that draws a histogram
ui <- fluidPage(

    # Application title
    titlePanel("Old Faithful Geyser Data"),

    # Sidebar with a slider input for number of bins 
    sidebarLayout(
        sidebarPanel(
            sliderInput(
            inputId = "bins",
            label = "Number of bins:",
            min = 1,
            max = 50,
            value = 30)
        ),

        # Show a plot of the generated distribution
        mainPanel(
           plotOutput("distPlot")
        )
    )
)

# Define server logic required to draw a histogram
server <- function(input, output) {

    output$distPlot <- renderPlot({
        # generate bins based on input$bins from ui.R
        x    <- faithful[, 2]
        bins <- seq(min(x), max(x), length.out = input$bins + 1)

        # draw the histogram with the specified number of bins
        hist(x, breaks = bins, col = 'darkgray', border = 'white',
             xlab = 'Waiting time to next eruption (in mins)',
             main = 'Histogram of waiting times')
    })
}

# Run the application 
shinyApp(ui = ui, server = server)

```

# Some examples

Using this structure, you can build complex dashboards with multiple pages and substantial computations in response to users' inputs:

https://agent-based-models.shinyapps.io/RegionalCOVIDSchoolSimulation/ (code: https://github.com/UCLA-PHP/school.epi.abm)

A particularly useful application is presenting interactive versions of paper figures and tables:

https://morrison.shinyapps.io/VaxEffApp/ (code: https://github.com/d-morrison/vax.eff)

Some more projects with public code:

https://ctsipopulationhealth.shinyapps.io/FluView/ (code: https://github.com/UCLA-PHP/FluView)

https://ctsipopulationhealth.shinyapps.io/wastewater/ (code: https://github.com/UCLA-PHP/wastewater)

# Interactive graphics with `plotly`

Use the `plotly` package to create interactive graphics; to start,  call `plotly::ggplotly()` on `ggplot2` graphs.

```{r}
library(ggplot2)
library(plotly)
library(maps)
data(canada.cities, package = "maps")

# create a graph in ggplot2:
viz <- ggplot(canada.cities, aes(long, lat)) +
  borders(regions = "canada") +
  coord_equal() +
  geom_point(aes(
    text = name, 
    size = pop), colour = "red", alpha = 1/2)

# convert to interactive via plotly:
plotly::ggplotly(viz)

```

You can pan, zoom, stretch and compress!

# Some key functions

## UI-side functions

### Input widgets

#### `sliderInput()`

Produces a slider:

```{r}
sliderInput(
  inputId = "slider", 
  label = "Text to go next to button",
  min = 0,
  max = 10,
  value = 5)

renderText(input$slider)
```

#### `numericInput()`

Produces a text box to input a number:

```{r}
numericInput(
  inputId = "how-I-want-to-refer-to-this-input-in-computations",
  label = "Give me a number!",
  min = -1,
  max = 4,
  value = 3.1415,
  step = 0.0001
  
)
renderText(input$"how-I-want-to-refer-to-this-input-in-computations")
```

#### `selectInput()`

Select from a list of options:

```{r}

selectInput("state", "What's your favourite state?", state.name)
renderText(input$state)
```

#### `radioButtons()`

```{r}
animals <- c("dog", "cat", "mouse", "bird", "other")
radioButtons(inputId = "animal", label = "What's your favourite animal?", animals)
renderText(input$animal)
```

#### `textInput()`

Free-text entry:

```{r}
textInput(inputId = "fav_color", label = "What's your favorite color?")
renderText(input$fav_color)
```

### Outputs

#### `renderTable()`/`tableOutput()`

```{r}
numericInput("nrows", "How many rows?", min = 1, max = nrow(iris), step = 1, value = 5)
renderTable(iris[1:input$nrows, ])
```

In UI-server apps, pair with an `Output()` function:

```r
# in server:
output$table1 = renderTable(iris[1:5, ])

# in UI:
tableOutput("table1")

```

#### `renderPlot()`/`plotOutput()`

```{r}

numericInput("maxnum", "How many points?", min = 1, step = 1, value = 5)
renderPlot(plot(1:input$maxnum))
```

(same deal with `plotOutput()` in UI)

## Other server-side functions

### `reactive()`

Creates an intermediate variable that sits between inputs and outputs; to avoid code duplication, avoid repeated computations, and improve readability. For example, if we want to display the histogram in the previous example twice, we could rewrite the `server` function above as follows:

```r
server <- function(input, output) {

    graph1 = reactive({
        # generate bins based on input$bins from ui.R
        x    <- faithful[, 2]
        bins <- seq(min(x), max(x), length.out = input$bins + 1)

        # draw the histogram with the specified number of bins
        hist(x, breaks = bins, col = 'darkgray', border = 'white',
             xlab = 'Waiting time to next eruption (in mins)',
             main = 'Histogram of waiting times')
    })

    output$distPlot <- renderPlot(graph1()) # note that reactive variables get used like functions, with parentheses at the end.
    
    output$Plot2 <- renderPlot(graph1()) # here we pass graph1() to a different output variable, to be used somewhere else in the UI.
}

```

# Your turn!

Let's try it! Modify a markdown file to make it interactive; pick any input that you weren't sure about, make an input widget for it, and wrap the outputs with `render` functions.

# Share your apps online!

In RStudio: File > Publish

# Key references

https://mastering-shiny.org/ (Hadley Wickham, also leads ggplot2, tidyverse)

https://engineering-shiny.org/index.html (advanced shiny app engineering)
